---
title: "Appendix"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r package, message=FALSE}
library(knitr)
library(kableExtra)
library(gridExtra)
library(tree)
library(rsample)
library(party)
library(reldist)
library(skimr)
library(tidyverse)
```

# Data Cleaning
## Import Dataset
```{r import data}
# Household data
hh <- read.csv("hhpub19.csv", header=T)

# Person data
pp <- read.csv("pppub19.csv", header=T)

# state db
fips = read.csv("fipscodes.csv", header = T)
```


## Add State names to Household database
```{r add state names}
# matching state name and fips code
hh <- left_join(hh,fips,by="GESTFIPS")
head(hh$GESTFIPS)
head(hh$state_full)
```

## Matching household and people database by Household ID (H_SEQ and PH_SEQ) 

```{r matching hh and pp}
# matching key between person data and household
#PH_SEQ is the sequence number of family record in household
match_hh_per <- unique(pp$PH_SEQ)

# hh database we use
new_hh <- hh %>%  filter(H_SEQ %in% match_hh_per)

# pp database we use
# we use only householders
new_pp <- pp %>% filter(HHDREL==1)

# match hh and pp
df <- inner_join(new_hh, new_pp, by=c("H_SEQ" = "PH_SEQ"))
```


## Select all variables we use
```{r variable list} 
##Variables: 
#HTOTVAL: household income
#A_SEX: 1 = Male, 2 = Female
#A_AGE
#PRCITSHP: 1 = Native, born in US; 
          #2 = Native, born in PR or US outlying area 
               #/ Native, born abroad of US parent(s) 
               #/ Foreign born, US cit by naturalization;
          #3 = Foreign born, not a US citizen
#PRDTRACE: 1 = White only; 2 = Black only; 3 = Others
#PEHSPNON: Are you Spanish, Hispanic, or Latino?  1 = Yes 2 = No
#PENATVTY: Place of Birth 1 = U.S., 2 = Outside of U.S. 
#PEFNTVTY: Father's Place of Birth 1 = U.S., 2 = Outside of U.S. 
#PEMNTVTY: Mother's Place of Birth 1 = U.S., 2 = Outside of U.S. 
#GESTFIPS
#state
#state_full

df1 <- dplyr::select (df, 
                      HTOTVAL, A_SEX, A_AGE, PRCITSHP, PRDTRACE, PEHSPNON,
                      PENATVTY, PEFNTVTY, PEMNTVTY, GESTFIPS, state, state_full)
```


## Data Cleaning
```{r cleaning data 1}
#There is no NA data in this dataset.

#PRCITSHP:
for (i in 1:length(df1$PRCITSHP)) {
    if (df1$PRCITSHP[i] == 3 | df1$PRCITSHP[i] == 4){
      df1$PRCITSHP[i] = 2
    } 
    else if (new_pp$PRCITSHP[i] == 5){
      df1$PRCITSHP[i] = 3
    }
}

#PRDTRACE: 
for (i in 1:length(df1$PRDTRACE)) {
    if(!(df1$PRDTRACE[i] == 1 | df1$PRDTRACE[i] == 2)) {
      df1$PRDTRACE[i] = 3
    }
}

#PENATVTY: Place of Birth 1 = U.S., 2 = Outside of U.S.
df1$PENATVTY <- ifelse(df1$PENATVTY == 57, 1, 2)

##PEFNTVTY: Father's Place of Birth 1 = U.S., 2 = Outside of U.S.
df1$PEFNTVTY <- ifelse(df1$PEFNTVTY == 57, 1, 2)

#PEMNTVTY: Mother's Place of Birth 1 = U.S., 2 = Outside of U.S.
df1$PEMNTVTY <- ifelse(df1$PEMNTVTY == 57, 1, 2)
```

```{r cleaning data 2}
colnames(df1) <- c("HHincome","Sex","A_Age", "Citizenship","Race","Hispanic",
                   "Place_of_Birth","F's_Place_of_Birth","M's_Place_of_Birth",
                   "GESTFIPS","state","state_full")
df1$HHincome <- as.integer(df1$HHincome)/1000
df1$Sex <- as.factor(ifelse(df1$Sex==1,"Male","Female"))
df1$Citizenship <- as.factor(ifelse(df1$Citizenship==1,"citizen",
                   ifelse(df1$Citizenship==2,"citizen born outside","non-citizen")))
df1$Race <- as.factor(ifelse(df1$Race==1, "White",
            ifelse(df1$Race==2,"Black","Other")))
df1$Hispanic <- as.factor(ifelse(df1$Hispanic==1,"Yes","No"))
df1$Place_of_Birth <- as.factor(ifelse(df1$Place_of_Birth==1,"US","non-US"))
df1$`F's_Place_of_Birth` <- as.factor(ifelse(df1$`F's_Place_of_Birth`==1,"US","non-US"))
df1$`M's_Place_of_Birth` <- as.factor(ifelse(df1$`M's_Place_of_Birth`==1,"US","non-US"))
head(df1)
```


# Summary Statistics
```{r summary statistics state}
sum_table.whole <- df1 %>%
  summarize(state="Whole",
            n=n(),
            ave.income=round(mean(HHincome),1),
            `25%tile`=round(quantile(HHincome,0.25),1),
            `75%tile`=round(quantile(HHincome,0.75),1),
            `s.d.`=round(sd(HHincome),1),
            `Gini`=round(gini(HHincome),3),
            `White ratio`=round(mean(I(Race=="White"))*100,1),
            `Male ratio`=round(mean(I(Sex=="Male"))*100,1),
            `Hispanic ratio`=round(mean(I(Hispanic=="Yes"))*100,1))

sum_table.state <- df1 %>%
  group_by(state)%>%
  summarise(n=n(),
            ave.income=round(mean(HHincome),1),
            `25%tile`=round(quantile(HHincome,0.25),1),
            `75%tile`=round(quantile(HHincome,0.75),1),
            `s.d.`=round(sd(HHincome),1),
            `Gini`=round(gini(HHincome),3),
            `White ratio`=round(mean(I(Race=="White"))*100,1),
            `Male ratio`=round(mean(I(Sex=="Male"))*100,1),
            `Hispanic ratio`=round(mean(I(Hispanic=="Yes"))*100,1))

sum_table <- as.data.frame(rbind(sum_table.whole,sum_table.state))
sum_table$n <- formatC(sum_table$n, format="d", big.mark=',')
kable(sum_table, align=rep('c', 10),booktabs = T) 
colnames(sum_table)
write.csv(sum_table,"sum_table.csv")
```



# Tree for the whole US
```{r data for whole us}
df1.whole <- df1 %>%
  select(HHincome,Sex,Citizenship,Race,Hispanic,
         Place_of_Birth,`F's_Place_of_Birth`,`M's_Place_of_Birth`)
head(df1.whole)
```

## Split the df into training and test set
```{r whole training and test set}
# split the df into training and test set
set.seed(12345)
train <- sample(c(1:nrow(df1.whole)),nrow(df1.whole)*0.75,replace = FALSE) 
test <- c(1:nrow(df1.whole))[!(c(1:nrow(df1.whole)) %in% train)]
df_train <- df1.whole[train,]
df_test <- df1.whole[test,]
```


## Cross validation for deciding alpha. We search among potential alpha from 0.01 to 0.1.
```{r whole CV}
# Develop function for CV (get MSE)

holdout_results <- function(splits,alpha) {
  # Fit the model to the training set
  
  mod <- ctree(HHincome ~ .,
                data=analysis(splits),
                control = ctree_control(mincriterion=1-alpha,
                                        testtype="Bonferroni"))
  # Get MSE based on test set
  test <- assessment(splits)
  mse <- 
    mean((test$HHincome - predict(mod,newdata = test))^2)
    
  mse
}

# set potential alpha
set.seed(23456)
a <-  seq(0.01, 0.1, 0.005)
cv_result <- data.frame(alpha=a)

# repeat CV for each alpha
for (i in 1:length(a)) {

 ## 10 folded CV
cv10 <- vfold_cv(data = df_train,v = 10) %>%
  mutate(results = pmap(list(splits,a[i]),holdout_results)) %>%
  unnest(results)

cv_result$mse[i] <- mean(cv10$results)
}

# optimal alpha
opt_alpha <- cv_result$alpha[cv_result$mse==min(cv_result$mse)]
opt_alpha
```


## Draw tree
```{r whole estimate tree}
# Baseline case
citree <- ctree(HHincome ~ .,
                data=df_train,
                control = ctree_control(mincriterion=1-opt_alpha,
                                        testtype="Bonferroni"))

```

```{r whole plot, fig.width=20, fig.height=10}
# Plot
plot(citree, type="simple")
```


## MSE and Opportunity based Gini coefficient
```{r whole MSE}
# mse
mse <- mean((df_test$HHincome - predict(citree,newdata = df_test))^2)
mse
```

```{r whole opportunity based gini coefficient}
# opportunity based gini coefficient
gini_opp <- gini(predict(citree,newdata = df1.whole))
gini <- gini(df1.whole$HHincome)
kable(data.frame(gini.opp=gini_opp,gini=gini))
```


## Robustness check
## Conditional Inference Random Forest

```{r Conditional Inference Random Forest}
ciforest <- 
  partykit::cforest(HHincome ~ .,
          data=df_train,
	        control = partykit::ctree_control(alpha=opt_alpha,
                                  testtype="Bonferroni"))
```

```{r Conditional Inference Random Forest MSE}
#mse
memory.limit(size=10000) 
mse_forest <- mean((df_test$HHincome - 
                      predict(ciforest,newdata = df_test,type = "response"))^2) 
kable(data.frame(mse.tree=mse,mse.forest=mse_forest))
```


# State-level Analysis
```{r data for state level}
df1.state <- df1 %>%
  select(HHincome,Sex,Citizenship,Race,Hispanic,Place_of_Birth,
         `F's_Place_of_Birth`,`M's_Place_of_Birth`,state)
head(df1.state)
```

## Split the df into training and test set for each state
```{r state training and test set}
# Generate subsample for each state
state_list <- vector(mode = "list", length = 51)
state_name <- unique(df1.state$state)
for (i in 1:51) {
    state_list[[i]] <- 
      df1.state %>% filter(state == state_name[i]) %>%
      dplyr::select(-state)
}


# Split the df into training and test set for each state
train_length <- 51
test_length <- 51
train_list <- vector(mode = "list", length = 51)
test_list <- vector(mode = "list", length = 51)

set.seed(3000)
for (i in 1:train_length) {
  train_list[[i]]=sample(1:nrow(state_list[[i]]),
                        nrow(state_list[[i]])*0.75)
  
  test_list[[i]]=
    c(1:nrow(state_list[[i]]))[!(c(1:nrow(state_list[[i]])) %in% train_list[[i]])]
}
```

## Cross validation for deciding alpha for each state
```{r state CV}
# function fot set potential alpha
a <-  seq(0.01, 0.1, 0.005)

cv_alpha <- function(df_train){
 cv_result <- data.frame(alpha=a)
 
 ## repeat CV for each alpha
 for (i in 1:length(a)) {
 ## 10 folded CV
 cv10 <- vfold_cv(data = df_train,v = 10) %>%
  mutate(results = pmap(list(splits,a[i]),holdout_results)) %>%
  unnest(results)
 cv_result$mse[i] <- mean(cv10$results)
 }
 
# optimal alpha
opt_alpha.state <- cv_result$alpha[cv_result$mse==min(cv_result$mse)]
opt_alpha.state
}

# CV for all states
set.seed(1234)
opt_alpha.state <- data.frame(state_group=c(1:51),state=state_name)
for (i in 1:train_length) {
  
  train <- train_list[[i]]
  opt_alpha.state$alpha[i] <- cv_alpha(state_list[[i]][train, ])
}
```

## Using optimal alpha, estimate Tree of each state
```{r state estimate tree training set}
# using only training set
tree_list.train <- vector(mode = "list", length = 51)
for (i in 1:length(tree_list.train)) {
 
 train <- train_list[[i]]
 citree <- ctree(HHincome ~ .,
                data=state_list[[i]][train,],
                control = 
                  ctree_control(mincriterion=1-opt_alpha.state$alpha[i],
                                testtype="Bonferroni"))
 
 tree_list.train[[i]] <- citree
}
```


```{r state estimate tree whole sample}
# use whole sample
tree_list <- vector(mode = "list", length = 51)
for (i in 1:length(tree_list)) {
 
 citree <- ctree(HHincome ~ .,
                data=state_list[[i]],
                control = 
                  ctree_control(mincriterion=1-opt_alpha.state$alpha[i],
                                testtype="Bonferroni"))
 
 tree_list[[i]] <- citree
}
```

## Draw Trees for sample states
```{r state draw tree, fig.width=10, fig.height=6}
plot(tree_list[[31]],type="simple") #KY
plot(tree_list[[7]],type="simple") #NY
plot(tree_list[[49]],type="simple") #CA
plot(tree_list[[45]],type="simple") #UT
plot(tree_list[[9]],type="simple") #PA
plot(tree_list[[34]],type="simple") #MS
plot(tree_list[[38]],type="simple") #TX
plot(tree_list[[29]],type="simple") #GA
plot(tree_list[[33]],type="simple") #AL
plot(tree_list[[4]],type="simple") #MA
plot(tree_list[[24]],type="simple") #DC
plot(tree_list[[12]],type="simple") #IL
plot(tree_list[[14]],type="simple") #WI
plot(tree_list[[47]],type="simple") #WA
plot(tree_list[[10]],type="simple") #OH
plot(tree_list[[8]],type="simple") #NJ
plot(tree_list[[44]],type="simple") #AZ
```

## MSE and Opportunity based Gini coefficient of each state
```{r state MSE}
# mse
mse.state <- c()
for (i in 1:51) {
  test <- test_list[[i]]
  mse.state[i] <- mean((state_list[[i]][test, ]$HHincome-
                    predict(tree_list.train[[i]],
                            newdata = state_list[[i]][test, ]))^2)
}
```

##Boosting and Bagging
```{r}
#Boosting
library(gbm)
set.seed(500)
boost_list.train <- vector(mode = "list", length = 51)
mse.boost <- rep(NA, 51)
for (i in 1:length(boost_list.train)) {
 
 train <- train_list[[i]]
 test <- test_list[[i]]
 boost.state <- gbm(HHincome ~ ., 
                    data=state_list[[i]][train,],
                    distribution="gaussian",
                    n.trees=1000,
                    shrinkage=opt_alpha.state$alpha[i],
                    interaction.depth = 4)
 boost_list.train[[i]] <- boost.state
 test.pred <- predict(boost.state, newdata=state_list[[i]][test,], n.trees=1000)
 mse.boost[i] <- mean((state_list[[i]][test,]$HHincome - test.pred)^2)
}

state <- data.frame(mse = mse.state)
boost <- data.frame(mse = mse.boost)

state$color <- 'state'
boost$color <- 'boost'

compare <- rbind(state, boost)

ggplot(compare, aes(mse, fill = color)) + geom_density(alpha = 0.2)

```


```{r state opportunity based gini coefficient}
# opportunity based gini coefficient
gini_opp.state <- c()
for (i in 1:51) {
  gini_opp.state[i] <- gini(predict(tree_list[[i]],newdata = state_list[[i]]))
}
# normal gini-coefficient
gini.state <- c()
for (i in 1:51) {
  gini.state[i] <- gini(state_list[[i]]$HHincome)
}
```

```{r gini summary state}
gini.my <- data.frame(state=state_name,
                      my.gini=round(gini.state,3),
                      my.gini.opp=round(gini_opp.state,3),
                      mse=round(mse.state,0),
                      opt_alpha=as.numeric(opt_alpha.state$alpha))
gini.official <- fips %>% select(state,gini)

gini.comp <- left_join(gini.my,gini.official,by="state")
kable(gini.comp,booktabs = T)
```


## State Mapping (opp base)
```{r State Mapping opp base}
library(usmap)
df.gini.opp <- data.frame(state=state_name, 
                      gini_opp=gini_opp.state)
df.gini.new <- df.gini.opp[order(-gini_opp.state),]
df.gini.new$order = c(1:51)
plot_usmap(data=df.gini.new, values="gini_opp", color='black', 
           labels = state_name,include=state_name)+
  scale_fill_continuous(
    low='white', high='red', name="opportunity base gini-coefficients")+
  theme(legend.position = 'right')
```

## Mapping (original base)
```{r State Mapping over all}
library(usmap)
df.gini <- data.frame(state=state_name, 
                      gini=gini.state)
plot_usmap(data=df.gini, values="gini", color='black',
           labels=state_name,include=state_name)+
  scale_fill_continuous(
    low='white', high='red', name="original gini-coefficients")+
  theme(legend.position = 'right')
```

## Mapping (official base)
```{r official gini}
library(usmap)
df.gini <- data.frame(state=state_name, 
                      gini=as.numeric(gini.comp$gini))
plot_usmap(data=df.gini, values="gini", color='black',
           labels=state_name,include=state_name)+
  scale_fill_continuous(
    low='white', high='red', name="original gini-coefficients")+
  theme(legend.position = 'right')
```


# Age-level Analysis
```{r data for age}
df1.age <- df1 %>%
  select(HHincome,Sex,Citizenship,Race,Hispanic,Place_of_Birth,
         `F's_Place_of_Birth`,`M's_Place_of_Birth`,A_Age)
head(df1.age)
```

## Generate age sub-group
```{r age sub-group}
# generate age flag
df1.age$AGE_flag[df1.age$A_Age>=18 & df1.age$A_Age<35]<-1
df1.age$AGE_flag[df1.age$A_Age>=35 & df1.age$A_Age<50]<-2
df1.age$AGE_flag[df1.age$A_Age>=50 & df1.age$A_Age<65]<-3
df1.age$AGE_flag[is.na(df1.age$AGE_flag)]<-0
```


## Summary Statistics (age)
```{r summary statistics age}
age.category <- c("18-34","35-49","50-64")

sum_table.age <- df1.age %>%
  group_by(AGE_flag)%>%
  summarise(sample.size=n(),
            ave.income=round(mean(HHincome),1),
            `25%tile`=round(quantile(HHincome,0.25),1),
            `75%tile`=round(quantile(HHincome,0.75),1),
            std.dev=round(sd(HHincome),1),
            Gini=round(gini(HHincome),3),
            male=round(mean(I(Sex=="Male"))*100,1),
            white=round(mean(I(Race=="White"))*100,1)) %>%
  filter(AGE_flag>0) %>%
  as.data.frame()


sum_table.age[,1] <- age.category

kable(sum_table.age, booktabs = T)
```


## Split the df into training and test set for each age group
```{r age training and test set}
# generate age sub-group dataset
age_list <- vector(mode = "list", length = 3)
for (i in 1:3) {
    age_list[[i]] <- df1.age %>% filter(AGE_flag == i) %>% select(-c(A_Age,AGE_flag))
}

# Generate Training and Testing Set List. We assign 75 % of sample to Training.
train_length.age <- 3
test_length.age <- 3
train_list.age <- vector(mode = "list", length = 3)
test_list.age <- vector(mode = "list", length = 3)
set.seed(5000)
for (i in 1:train_length.age) {
  train_list.age[[i]]=sample(1:nrow(age_list[[i]]),
                        nrow(age_list[[i]])*0.75)
  
  test_list.age[[i]]=
    c(1:nrow(age_list[[i]]))[!(c(1:nrow(age_list[[i]])) %in% train_list.age[[i]])]
}
```

## Conduct CV to set optimal alpha
```{r age CV}
# CV for all age group
set.seed(12345)
opt_alpha.age <- data.frame(age_group=c(1:3))

for (i in 1:train_length.age) {
  
  train <- train_list.age[[i]]
  opt_alpha.age$alpha[i] <- cv_alpha(age_list[[i]][train, ])
}
```

## Using optimal alpha, estimate Tree of each age group
```{r age estimate tree training set}
# using only training set
tree_list.age.train <- vector(mode = "list", length = 3)
for (i in 1:length(tree_list.age.train)) {
 
 train <- train_list.age[[i]]
 citree <- ctree(HHincome ~ .,
                data=age_list[[i]][train,],
                control = 
                  ctree_control(mincriterion=1-opt_alpha.age$alpha[i],
                                testtype="Bonferroni"))
 
 tree_list.age.train[[i]] <- citree
}
```


```{r age estimate tree whole sample}
# use whole sample
tree_list.age <- vector(mode = "list", length = 3)
for (i in 1:length(tree_list.age)) {
 train <- train_list.age[[i]]
 citree <- ctree(HHincome ~ .,
                data=age_list[[i]],
                control = ctree_control(mincriterion=1-opt_alpha.age$alpha[i],
                                        testtype="Bonferroni"))
 tree_list.age[[i]] <- citree
}
```


## Draw Tree of each age group
```{r age group plot, fig.width=13, fig.height=10}
plot(tree_list.age[[1]],type="simple") 
plot(tree_list.age[[2]],type="simple")
plot(tree_list.age[[3]],type="simple") 
```

## MSE and Opportunity based Gini coefficient for 5 age groups
```{r age MSE}
# mse
mse.age <- c()
for (i in 1:3) {
  test <- test_list.age[[i]]
  mse.age[i] <- mean((age_list[[i]][test, ]$HHincome-
                    predict(tree_list.age.train[[i]],
                            newdata = age_list[[i]][test, ]))^2)
}
```



```{r age opportunity based gini coefficient}
# opportunity based gini coefficient
gini_opp.age <- c()
for (i in 1:3) {
  gini_opp.age[i] <- gini(predict(tree_list.age[[i]],newdata = age_list[[i]]))
}
```


```{r gini summary age}
gini.comp.age <- data.frame(age=age.category,
                      gini=round(as.numeric(sum_table.age$Gini),3),
                      my.gini.opp=round(gini_opp.age,3),
                      mse=round(mse.age,0),
                      opt_alpha=as.numeric(opt_alpha.age$alpha))

kable(gini.comp.age,booktabs = T)
```


