---
title: "age_tree"
author: "Takahiro Minami"
date: "2/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(mc.cores = parallel::detectCores())
```

```{r setup, include=FALSE}
library(gridExtra)
library(broom)
library(modelr)
library(car)
library(knitr)
library(MASS)
library(foreign)
library(ISLR)
library(rsample)
library(yardstick)
library(tree)
library(e1071)
library(class) # for kNN
library(arm) # for a few base features
library(partykit)
library(reldist) # gini coefficient
library(skimr)
library(tidyverse)
options(width=70, digits=4, scipen=8)
```


##Loading Data
```{r}
df1 <- read.csv("MLData_Clean.csv")
```

```{r} 
##Variables: 
#HTOTVAL: household income
#H_TENURE
#A_SEX: 1 = Male, 2 = Female
#A_AGE
#PRCITSHP: 1 = Native, born in US; 2 = Native, born in PR or US outlying area / Native, born abroad of US parent(s) / Foreign born, US cit by naturalization; 3 = Foreign born, not a US citizen
#PRDTRACE: 1 = White only; 2 = Black only; 3 = American Indian, Alaskan Native only (AI); 4 = Others
#PEHSPNON: Are you Spanish, Hispanic, or Latino?  1 = Yes 2 = No
#PENATVTY: Place of Birth 1 = U.S., 2 = Outside of U.S. 
#PEFNTVTY: Father's Place of Birth 1 = U.S., 2 = Outside of U.S. 
#PEMNTVTY: Mother's Place of Birth 1 = U.S., 2 = Outside of U.S. 
#A_MJOCC: Major Occupation Code 0-11, Merge 6 and 11 to 0 
#A_MJIND: Major Industry Code 0-14, merge 14 to 0, merge 7 to 12, merge 1 and 2
#GESTFIPS
#state
#state_full
```


## We need factorize variables
```{r}
# clean data
df1[sapply(df1, is.integer)] <- lapply(df1[sapply(df1, is.integer)], as.factor)
df1$HTOTVAL <- as.integer(df1$HTOTVAL)/1000
df1$A_AGE <- as.integer(df1$A_AGE)
df1 <- select(df1,-c(X,state,state_full,GESTFIPS))
head(df1)
```

## Generate age groups. 

### Group 1: 18-24
### Group 2: 25-34
### Group 3: 35-44
### Group 4: 45-54
### Group 5: 55-64
```{r}
# generate age flag
df1$AGE_flag[df1$A_AGE>=18 & df1$A_AGE<25]<-1
df1$AGE_flag[df1$A_AGE>=25 & df1$A_AGE<35]<-2
df1$AGE_flag[df1$A_AGE>=35 & df1$A_AGE<45]<-3
df1$AGE_flag[df1$A_AGE>=45 & df1$A_AGE<55]<-4
df1$AGE_flag[df1$A_AGE>=55 & df1$A_AGE<65]<-5
df1$AGE_flag[is.na(df1$AGE_flag)]<-0
```

```{r}
age_list <- vector(mode = "list", length = 5)
for (i in 1:5) {
    age_list[[i]] <- df1 %>% filter(AGE_flag == i) %>% select(-c(A_AGE,AGE_flag))
}
```

## Summary of each age groups
```{r}
df1 %>% filter(AGE_flag>0) %>% group_by(AGE_flag) %>%
  summarise(n=n(),
            one_quarter_income=quantile(HTOTVAL,0.25),
            median_income=quantile(HTOTVAL,0.5),
            three_quarter_income=quantile(HTOTVAL,0.75))%>%
  kable()
```

## Generate Training and Testing Set List. We assign 75 % of sample to Training.
```{r}
train_length <- 5
test_length <- 5
train_list <- vector(mode = "list", length = 5)
test_list <- vector(mode = "list", length = 5)
set.seed(100)
for (i in 1:train_length) {
  train_list[[i]]=sample(1:nrow(age_list[[i]]),
                        nrow(age_list[[i]])*0.75)
  
  test_list[[i]]=
    c(1:nrow(age_list[[i]]))[!(c(1:nrow(age_list[[i]])) %in% train_list[[i]])]
}
```


## Conduct CV to set optimal alpha. We search among potential alpha from 0.01 to 0.1.
```{r}
# Develop function for CV (get MSE)

holdout_results <- function(splits,alpha) {
  # Fit the model to the training set
  mod <- ctree(HTOTVAL ~ .,
                data=analysis(splits),
                control = ctree_control(alpha=alpha,
                                        testtype="Bonferroni"))
  # Get MSE based on test set
  test <- assessment(splits)
  mse <- 
    mean((test$HTOTVAL - predict(mod,newdata = test))^2)
    
  mse
}

# function fot set potential alpha
a <-  seq(0.01, 0.1, 0.005)
cv_alpha <- function(df_train){

cv_result <- data.frame(alpha=a)

# repeat CV for each alpha
for (i in 1:length(a)) {

 ## 10 folded CV
cv10 <- vfold_cv(data = df_train,v = 10) %>%
  mutate(results = pmap(list(splits,a[i]),holdout_results)) %>%
  unnest(results)

cv_result$mes[i] <- mean(cv10$results)
}

# optimal alpha
opt_alpha <- cv_result$alpha[cv_result$mes==min(cv_result$mes)]
opt_alpha
}
```


```{r}
# CV for all age group
set.seed(1234)
opt_alpha <- data.frame(age_group=c(1:5))

for (i in 1:train_length) {
  
  train <- train_list[[i]]
  opt_alpha$alpha[i] <- cv_alpha(age_list[[i]][train, ])
}

```



## Using optimal alpha, Draw Tree of each age group
```{r}
tree_list <- vector(mode = "list", length = 5)
for (i in 1:length(tree_list)) {
 train <- train_list[[i]]
 citree <- ctree(HTOTVAL ~ .,
                data=age_list[[i]][train, ],
                control = ctree_control(alpha=opt_alpha$alpha[i],
                                        testtype="Bonferroni"))
 tree_list[[i]] <- citree
}
```


## Graph
```{r, fig.width=50, fig.height=20}
plot(tree_list[[1]]) 
plot(tree_list[[2]])
plot(tree_list[[3]])
plot(tree_list[[4]])
plot(tree_list[[5]])
```


## MES and Opportunity based Gini coefficient of 5 age groups
```{r}
# mse
mse <- c()
for (i in 1:5) {
  test <- test_list[[i]]
  mse[i] <- mean((age_list[[i]][test, ]$HTOTVAL-
                    predict(tree_list[[i]],
                            newdata = age_list[[i]][test, ]))^2)
}





# opportunity based gini coefficient
gini_opp <- seq(from=1, to=5, by=1)
for (i in 1:5) {
  gini_opp[i] <- gini(predict(tree_list[[i]],newdata = age_list[[i]]))
}
```

