---
title: "age_tree"
author: "Takahiro Minami"
date: "2/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(mc.cores = parallel::detectCores())
```

```{r setup, include=FALSE}
library(gridExtra)
library(broom)
library(modelr)
library(car)
library(knitr)
library(MASS)
library(foreign)
library(ISLR)
library(rsample)
library(yardstick)
library(tree)
library(e1071)
library(class) # for kNN
library(arm) # for a few base features
library(partykit)
library(reldist) # gini coefficient
library(skimr)
library(tidyverse)
options(width=70, digits=4, scipen=8)
```


##Loading Data
```{r}
df1 <- read.csv("MLData_Clean.csv")
```

```{r} 
##Variables: 
#HTOTVAL: household income
#H_TENURE
#A_SEX: 1 = Male, 2 = Female
#A_AGE
#PRCITSHP: 1 = Native, born in US; 2 = Native, born in PR or US outlying area / Native, born abroad of US parent(s) / Foreign born, US cit by naturalization; 3 = Foreign born, not a US citizen
#PRDTRACE: 1 = White only; 2 = Black only; 3 = American Indian, Alaskan Native only (AI); 4 = Others
#PEHSPNON: Are you Spanish, Hispanic, or Latino?  1 = Yes 2 = No
#PENATVTY: Place of Birth 1 = U.S., 2 = Outside of U.S. 
#PEFNTVTY: Father's Place of Birth 1 = U.S., 2 = Outside of U.S. 
#PEMNTVTY: Mother's Place of Birth 1 = U.S., 2 = Outside of U.S. 
#A_MJOCC: Major Occupation Code 0-11, Merge 6 and 11 to 0 
#A_MJIND: Major Industry Code 0-14, merge 14 to 0, merge 7 to 12, merge 1 and 2
#GESTFIPS
#state
#state_full
```


## We need factorize variables
## We only use sample over 18 years old
```{r}
# clean data
df1$H_TENURE[df1$H_TENURE==3] <- 2
df1[sapply(df1, is.integer)] <- lapply(df1[sapply(df1, is.integer)], as.factor)
df1$HTOTVAL <- as.integer(df1$HTOTVAL)/1000
df1$A_AGE <- as.integer(df1$A_AGE)
df1$H_TENURE <- as.factor(df1$H_TENURE)
df1 <- select(df1,-X) %>% filter(A_AGE>=18)%>% select(-A_AGE,A_MJOCC,A_MJIND,H_TENURE)
head(df1)
```


```{r}
state_list <- vector(mode = "list", length = 51)
state_name <- unique(df1$state)
for (i in 1:51) {
    state_list[[i]] <- df1 %>% filter(state == state_name[i]) %>% select(-c(GESTFIPS,state))
}
```

## Summary of each age groups
```{r}
df1 %>% group_by(state_full) %>%
  summarise(n=n(),
            one_quarter_income=quantile(HTOTVAL,0.25),
            median_income=quantile(HTOTVAL,0.5),
            three_quarter_income=quantile(HTOTVAL,0.75))%>%
  kable()
df1 <- df1 %>% select(-state_full)
```

## Generate Training and Testing Set List. We assign 75 % of sample to Training.
```{r}
train_length <- 51
test_length <- 51
train_list <- vector(mode = "list", length = 51)
test_list <- vector(mode = "list", length = 51)
set.seed(2000)
for (i in 1:train_length) {
  train_list[[i]]=sample(1:nrow(state_list[[i]]),
                        nrow(state_list[[i]])*0.75)
  
  test_list[[i]]=
    c(1:nrow(state_list[[i]]))[!(c(1:nrow(state_list[[i]])) %in% train_list[[i]])]
}
```


## Conduct CV to set optimal alpha. We search among potential alpha from 0.01 to 0.1.
```{r}
# Develop function for CV (get MSE)

holdout_results <- function(splits,alpha) {
  # Fit the model to the training set
  mod <- ctree(HTOTVAL ~ .,
                data=analysis(splits),
                control = ctree_control(alpha=alpha,
                                        testtype="Bonferroni"))
  # Get MSE based on test set
  test <- assessment(splits)
  mse <- 
    mean((test$HTOTVAL - predict(mod,newdata = test))^2)
    
  mse
}

# function fot set potential alpha
a <-  seq(0.01, 0.1, 0.005)
cv_alpha <- function(df_train){

cv_result <- data.frame(alpha=a)

# repeat CV for each alpha
for (i in 1:length(a)) {

 ## 10 folded CV
cv10 <- vfold_cv(data = df_train,v = 10) %>%
  mutate(results = pmap(list(splits,a[i]),holdout_results)) %>%
  unnest(results)

cv_result$mes[i] <- mean(cv10$results)
}

# optimal alpha
opt_alpha <- cv_result$alpha[cv_result$mes==min(cv_result$mes)]
opt_alpha
}
```


```{r}
# CV for all states
set.seed(1234)
opt_alpha <- data.frame(state_group=c(1:51),state=state_name)

for (i in 1:train_length) {
  
  train <- train_list[[i]]
  opt_alpha$alpha[i] <- cv_alpha(state_list[[i]][train, ])
}

```



## Using optimal alpha, Draw Tree of each state
```{r}
tree_list <- vector(mode = "list", length = 51)
for (i in 1:length(tree_list)) {
 train <- train_list[[i]]
 citree <- ctree(HTOTVAL ~ .,
                data=state_list[[i]][train, ],
                control = ctree_control(alpha=0.03,
                                        testtype="Bonferroni"))
 tree_list[[i]] <- citree
}
```


## Graph
```{r, fig.width=50, fig.height=20}
plot(tree_list[[1]]) 
plot(tree_list[[2]])
plot(tree_list[[3]])
plot(tree_list[[4]])
plot(tree_list[[5]])
```


## MES and Opportunity based Gini coefficient of 5 age groups
```{r}
# mse
mse <- c()
for (i in 1:51) {
  test <- test_list[[i]]
  mse[i] <- mean((state_list[[i]][test, ]$HTOTVAL-
                    predict(tree_list[[i]],
                            newdata = state_list[[i]][test, ]))^2)
}





# opportunity based gini coefficient
gini_opp <- c()
for (i in 1:51) {
  gini_opp[i] <- gini(predict(tree_list[[i]],newdata = state_list[[i]]))
}

# normal gini-coefficient
gini <- c()
for (i in 1:51) {
  gini[i] <- gini(state_list[[i]]$HTOTVAL)
}
```


## Mapping (opp base)
```{r}
library(usmap)

df.gini.opp <- data.frame(state=state_name, 
                      gini_opp=gini_opp)

plot_usmap(data=df.gini.opp, values="gini_opp", color='black',
           include=state_name)+
  scale_fill_continuous(
    low='gray', high='red', name="opportunity base gini-coefficients")+
  theme(legend.position = 'right')
```

## Mapping (original base)
```{r}
library(usmap)

df.gini.2018 <- read.csv("fipscodes.csv")

df.gini <- df.gini.2018%>%select(state,gini)

plot_usmap(data=df.gini, values="gini", color='black',
           include=state_name)+
  scale_fill_continuous(
    low='gray', high='red', name="original gini-coefficients")+
  theme(legend.position = 'right')
```

## Mapping (original base - opp base)
```{r}
library(usmap)

df.gini.eff <- data.frame(state=state_name, 
                      gini_eff=gini-gini_opp)

plot_usmap(data=df.gini.eff, values="gini_eff", color='black',
           include=state_name)+
  scale_fill_continuous(
    low='white', high='red', name="priginal - opportunity base gini-coefficients")+
  theme(legend.position = 'right')
```

